---
title: "cyclist_project_case_study"
author: "Greg"
date: "2023-11-21"
output: html_document
---

#Project Capstone Cyclist

##Step 1 Spreadsheet (Excel)

* Download all data files (may-oct 2023)
* Create copies to transform to xlsx documents (leave csv backup copies)
* Added ride_length (ended_at - started_at)(D2 - C2)
* Added day_of_week (WEEKDAY(started_at)) (WEEKDAY(C2)) as number
* Exported all the files to transform in RStudio

##* Complete raw dataset is hosted on Kaggle.com [click here](https://www.kaggle.com/datasets/gregoriolobo/cyclist-complete-dataset-may-oct)

##Step 2 RStudio

* Prepare environment
```{r prepare environment, echo=FALSE}
library(readxl)
library(skimr)
library(tidyverse)
```

* Assign all files to individual dataframes numbered 5-10 (may-oct)

```{r create and assign dataframes, echo=FALSE}
df_05 <- read_excel("G:/My Drive/Google Data Analytics Certificate/Module 8 Capstone/2023-cyclist_xlsx_files/202305-divvy-tripdata.xlsx")
df_06 <- read_excel("G:/My Drive/Google Data Analytics Certificate/Module 8 Capstone/2023-cyclist_xlsx_files/202306-divvy-tripdata.xlsx")
df_07 <- read_excel("G:/My Drive/Google Data Analytics Certificate/Module 8 Capstone/2023-cyclist_xlsx_files/202307-divvy-tripdata.xlsx")
df_08 <- read_excel("G:/My Drive/Google Data Analytics Certificate/Module 8 Capstone/2023-cyclist_xlsx_files/202308-divvy-tripdata.xlsx")
df_09 <- read_excel("G:/My Drive/Google Data Analytics Certificate/Module 8 Capstone/2023-cyclist_xlsx_files/202309-divvy-tripdata.xlsx")
df_10 <- read_excel("G:/My Drive/Google Data Analytics Certificate/Module 8 Capstone/2023-cyclist_xlsx_files/202310-divvy-tripdata.xlsx")
```

* Merge all dataframes into a single df_complete
* First df structure evaluation

```{r merge dataframe, echo=TRUE}
df_complete <- rbind(df_05, df_06, df_07, df_08, df_09, df_10)
skim_without_charts(df_complete)
```

```{r initial output file, eval=FALSE, include=FALSE}
write_csv(df_complete, "G:/My Drive/Google Data Analytics Certificate/Module 8 Capstone/2023-cyclist_xlsx_files/2023-complete-tripdata.csv")
```

* Notice how started_at, ended_at and ride_length are of type POSIXct
* Change column ride_length to be calculated in seconds and as a numeric type to run calculations

```{r convert ride_length column, echo=FALSE}
df_complete$ride_length <- difftime(df_complete$ended_at, df_complete$started_at)
df_complete$ride_length <- as.numeric(as.character(df_complete$ride_length))
```

* DF structure summary and evaluation

```{r second df evaluation}
skim_without_charts(df_complete)
glimpse(df_complete)
str(df_complete)
colnames(df_complete)
```

* Verify columns member_casual and rideable_type to make sure all the values are correct

```{r column verification}
table(df_complete$member_casual)
table(df_complete$rideable_type)
```

* Create new aggregation columns as date, month, day, and day_of_week to make further calculations

```{r new aggregation columns, echo=FALSE}
df_complete$date <- as.Date(df_complete$started_at)
df_complete$month <- format(as.Date(df_complete$started_at), format = "%m")
df_complete$day <- format(as.Date(df_complete$started_at), format = "%d")
df_complete$day_of_week <- format(as.Date(df_complete$started_at), "%A")
```
 
* Create a new v2 df with cleaned data and removed negative erroneous values in ride_length column (167 records removed)

```{r v2 df with cleaned records}
df_complete_v2 <- df_complete[!df_complete$ride_length<0,]
```

* Descriptive analysis of ride_length column

```{r v2 ride_length summary}
summary(df_complete_v2$ride_length)
```

* Comparative statistics between members and casual riders

```{r members vs casual comparison}
aggregate(df_complete_v2$ride_length ~ df_complete_v2$member_casual, FUN = mean)
aggregate(df_complete_v2$ride_length ~ df_complete_v2$member_casual, FUN = median)
aggregate(df_complete_v2$ride_length ~ df_complete_v2$member_casual, FUN = max)
aggregate(df_complete_v2$ride_length ~ df_complete_v2$member_casual, FUN = min)
```


* Comparative average ride time by each day for members vs casual users

```{r average members vs casual per weekday}
aggregate(df_complete_v2$ride_length ~ df_complete_v2$member_casual + df_complete_v2$day_of_week, FUN = mean)
```

* Notice weekdays are in alphabetical order, change them in sequential order

```{r order weekdays, echo=FALSE}
df_complete_v2$day_of_week <- ordered(df_complete_v2$day_of_week, levels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

* Comparative average ride time by each day for members vs casual users
```{r average ride time members vs casual}
aggregate(df_complete_v2$ride_length ~ df_complete_v2$member_casual + df_complete_v2$day_of_week, FUN = mean)
```

* Analyze ridership data by type and weekday
```{r number of rides by type and weekday}
df_complete_v2 %>% 
  group_by(member_casual, day_of_week) %>%                #groups by usertype and weekday
  summarise(number_of_rides = n()							            #calculates the number of rides and average duration 
            ,average_duration = mean(ride_length)) %>% 		#calculates the average duration
  arrange(member_casual, day_of_week)
```

* Viz the number of rides by rider type and weekday
```{r viz number of rides per rider type}
df_complete_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) + 
  geom_col(position = "dodge")
```

* Viz for average duration by rider type and weekday
```{r viz average duration}
df_complete_v2 %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, day_of_week)  %>% 
  ggplot(aes(x = day_of_week, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```

* Export file for further analysis
```{r output file, eval=FALSE, include=FALSE}
write_csv(df_complete_v2, "G:/My Drive/Google Data Analytics Certificate/Module 8 Capstone/2023-cyclist_xlsx_files/2023-cleaned-final-trip-dataset.csv")
```